Microlensing as a possible probe of event-horizon structure in quasars

### 2016-07-19 paper1/ms.tex
editing

Section~\ref{sec:fold-crossing} shows the light curves that result
when each of the model sources crosses an ideal fold.  This would
apply in Figure~\ref{fig:magnification_map} to sources along the path
AB or BC, for sources small enough that the curvature of the caustics
is negligible.  With this assumption one can imagine the caustic as an
infinite wall to be crossed by the source as presented in
Figure~\ref{fig:infinite_fold}.  The source brightness distribution
parallel to the caustic naturally makes no difference to the
observable brightness; each source can be replaced by an effective
one-dimensional source profile, by flattening the source so it becomes
perpendicular to the caustic.  In principle, the effective
one-dimensional brightness profile could be recovered from the light
curve by deconvolution.  \cite{1999ApJ...524...49A} modelled this
profile as the result of a circular accretion disc seen through the
spacetime around a Kerr black hole, and \cite{2012MNRAS.423..676A}
have applied the idea to observed light curves to infer properties of
quasar accretion discs.  In this work, we take a simpler but arguably
more robust approach: we study features in the light curves
characteristic of a crescent-like source which in turn would indicate
a black-hole silhouette.  Figure~\ref{fig:char_points} shows the
qualitative features: there is a period during which the dark cutout
disc is crossing the caustic, and before and after there are periods
when the only the bright parts of the crescent are in transit across
the caustic.  The details depend on the orientation of the crescent,
but basically the dark disc causes a rising light curve to plateau or
dip.  These features are still present, albeit faintly, if the simple
crescent is replaced by a source based on an accretion-disc simulation
of a black-hole environment
(Figures~\ref{fig:M87_image}--\ref{fig:M87_plots}).

@@

We considered two sources, as follows.
\begin{itemize}
\item The crescent source (as explained in \S\ref{subsec:crescent})
  with parameters
\begin{equation}
   (R_p, R_n/Rp, a/Rp, b/Rp) = (1.31, 0.4, 0, 0.3)
\label{eqn:cp}
\end{equation}
corresponding to $r_{1/2}=1$.
\item A Gaussian source (as explained in \S\ref{subsec:crescent}),
  also with $r_{1/2}=1$.
\end{itemize}
Unit length has 30 pixels on the magnification map and 30 points on a
light curve.

Each lightcurve has 251 points regularly spaced in time, with Gaussian
noise at the level of 4\% of the current brightness.

Mock data were generated by taking path-I lightcurves for

Each mock light curve also has three nuisance parameters, namely the
beginning and end of the event and the brightness normalisation.
These are to be marginalised out by the MCMC.

@@

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/cc_forward.eps}
\caption{\label{fig:cc_forward} $\chi^2=243, P=0.32$.}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/cc_backward.eps}
\caption{\label{fig:cc_backward} $\chi^2=284, P=0.01$.}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/gc_forward.eps}
\caption{\label{fig:gc_forward} $\chi^2=344, P=3\times10^{-6}$.}
\end{figure}

\subsection{Model-fitting results}

Each of the three mock light curves was fitted to all three models.
We denote the nine possible cases with letter pairs, with the first
letter denoting the assumed model for the fitting procedure and 
the second letter representing
the source: thus CG means a {\em C\/}rescent model was fitted to mock
data from a {\em G\/}aussian source, DC stands for a uniform {\em
  D\/}isc model fitted to mock data from a {\em C\/}rescent source,
and so on.

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/Rp4all.eps}
\caption{\label{fig:mcmc} Posterior probability distribution for the
  source size $R_p$.  The vertical line is the correct value. Same
  color represents the same dataset whereas same line style
  corresponds to same model fit.  The legend gives the reduced
  $\chi^2$ of the best fit in each case.}
\end{figure}

Figure~\ref{fig:mcmc} shows the posterior probability distribution of
$R_p$ for all nine cases, along with the minimum reduced $\chi^2$ for
each case.  The area under each curve is unity. Note that the height
of the curves are not the likelihood, they are probability densities
in parameter space.

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/Rhalf_RnRp.eps}
  \includegraphics[width=0.9\hsize]{figures/aRp_bRp.eps}
\caption{\label{fig:crescentfit} The 2$\sigma$ contours (or error ellipses)
  of the crescent model when fitting with the three different
  datasets.}
\end{figure}

Let us first consider the three cases where a crescent model was
fitted.  These are the solid curves in Figure~\ref{fig:mcmc}, with the
colours of the curves indicating the source.  Meanwhile,
Figure~\ref{fig:crescentfit} shows $2\sigma$ of the inferred parameter
values.

\begin{enumerate}

\item[1) {\bf CC}:] solid red curve in Figure~\ref{fig:mcmc} and red
  ellipses in Figure~\ref{fig:crescentfit}.  In this case, a light
  curve from a crescent source was being fitted to a crescent model.
  The fit gives reduced $\chi^2$ close to unity, as expected.  The
  recovered parameter values are near or slightly outside the
  $2\sigma$ ellipses.  This is expected since we added a small
  systematic error, by using different caustics (though both clean
  folds) for the mock data and the model fit.  The apparent degeneracy
  between $a$ and $b$, seen in Figure~\ref{fig:crescentfit}, is also
  expected, since only the distance of the crescent's small circle
  from the caustic influences the magnification.

\item[2) {\bf CD}:] solid blue curve and blue ellipses.  Here a
  crescent model was fitted to a light curve from a uniform disc.  The
  situation is formally a crescent with $R_n=0$ and $a,b$ arbitrary,
  and this shows in the blue ellipses in the recovered parameter
  values.  The redundant parameters $\alpha$ and $\beta$, in effect, allow the
  model to partly fit the noise, and hence, the $\chi^2$ is somewhat
  lower than in the previous case.

\begin{figure}
\centering
\includegraphics[width=0.9\hsize]{figures/burnin_cg.eps}
\caption{\label{fig:burnin} The $\chi^2$ of each step of the full MCMC
  for one case.}
\end{figure}

\item[3) {\bf CG}:] Solid green curve and green ellipses.
  Additionally, Figure~\ref{fig:burnin} shows the progress of
  reduction of $\chi^2$ in this case, where a crescent-source model is
  used to fit data from a Gaussian source.  The parameter values are
  seemingly tightly constrained, but the recovered $R_p$ is completely
  wrong since the correct value is far from the green curve and green
  ellipses.  The reduced $\chi^2$ is only 0.85, indicative of
  over-fitting.  We can see what has happened from the red curve in
  the bottom panel of Figure~\ref{fig:mockdata}.  The best-fit model
  covers only a small part of the light curve.  That is, the fitting
  procedure has exploited the nuisance parameters to find a good fit
  to mainly noise.

\end{enumerate}

Next we consider the three cases where a disc model is fitted.  These
correspond to the dashed curves in Figure~\ref{fig:mcmc}.  There is
only one interesting parameter to fit, the disc radius $R_p$, and no
equivalent of Figure~\ref{fig:crescentfit} is needed.

\begin{enumerate}

\item[4) {\bf DC}:] red dashed curve.  On fitting a disc model to a
  crescent-source source, the recovered $R_p$ is incorrect, but the
  reduced $\chi^2$ of 1.29 signals that the data reject the model.

\item[5) {\bf DD}:] blue dashed curve.  When the correct model is
  fitted to a disc-source, the best fit $\chi^2=0.91$ is good and
  $R_p$ is recovered within the uncertainty estimate.

\item[6) {\bf DG}:] green dashed curve.  On fitting a disc model to a
  Gaussian-source model, the recovered $R_p$ is incorrect, but the
  reduced $\chi^2$ gives no signal that something is amiss.  It
  appears that a Gaussian source could mimic a uniform disc.

\end{enumerate}

Finally, we consider the three cases where a Gaussian model is fitted.
These correspond to the dotted curves in Figure~\ref{fig:mcmc}.  Again,
there is only one interesting parameter to fit, the Gaussian
$3\sigma$-radius, which we have called $R_p$.

\begin{enumerate}

\item[7) {\bf GC}:] red dotted curve.  When a Gaussian-source model is
  fitted to data from a crescent source, the recovered $R_p$ is wrong
  but the reduced $\chi^2=1.29$ shows the data rejecting the model.

\item[8) {\bf GD}:] blue dotted curve.  When a Gaussian-source model
  is fitted to data from a uniform disc, the recovered $R_p$ is wrong
  but the model is rejected anyway.

\item[9) {\bf GG}:] green dotted curve.  When data from a Gaussian is
  fitted to the correct model, $R_p$ is recovered within it's estimated
  uncertainty and the best fit reduced $\chi^2$ is close to unity.

\end{enumerate}

The above results suggest the following strategy for fitting a
lightcurve from an unknown source profile: first try a Gaussian-source
model; if the data reject that model, try a uniform disc; if the
uniform disc is also rejected by the data, try a crescent model.  Our
numerical experiments indicate --- assuming one of the three sources
models are correct --- that the reduced $\chi^2$ would unmask the
correct model, and its parameters would be correctly recovered.

>>>>

Section~\ref{sec:fold-crossing} shows the light curves that result
when each of the model sources crosses an ideal fold.  This would
apply in Figure~\ref{fig:magnification_map} to sources along the path
AB or BC, for sources small enough that the curvature of the caustics
is negligible.  With this assumption one can imagine the caustic as an
infinite wall to be crossed by the source as presented in
Figure~\ref{fig:infinite_fold}.  The source brightness distribution
parallel to the caustic naturally makes no difference to the
observable brightness; each source can be replaced by an effective
one-dimensional source profile, by flattening the source so it becomes
perpendicular to the caustic.  In principle, the effective
one-dimensional brightness profile could be recovered from the light
curve by deconvolution.  \cite{1999ApJ...524...49A} modelled this
profile as the result of a circular accretion disc seen through the
spacetime around a Kerr black hole, and \cite{2012MNRAS.423..676A}
have applied the idea to observed light curves to infer properties of
quasar accretion discs.  In this work, we take a simpler but arguably
more robust approach: we study features in the light curves
characteristic of a crescent-like source which in turn would indicate
a black-hole silhouette.  Figure~\ref{fig:char_points} shows the
qualitative features: there is a period during which the dark cutout
disc is crossing the caustic, and before and after there are periods
when the only the bright parts of the crescent are in transit across
the caustic.  The details depend on the orientation of the crescent,
but basically the dark disc causes a rising light curve to plateau or
dip.  These features are still present, albeit faintly, if the simple
crescent is replaced by a source based on an accretion-disc simulation
of a black-hole environment
(Figure~\ref{fig:M87_plots}).

@@

The mock lightcurve used a crescent source (as explained in
\S\ref{subsec:crescent}) with parameters
\begin{equation}
   \left(R_p, \frac{R_n}{R_p}, \frac{a}{R_p}, \frac{b}{R_p}\right) =
   (1.31, 0.4, 0, 0.3)
\label{eqn:cp}
\end{equation}
corresponding to $r_{1/2}=1$.  Unit length has 30 pixels on the
magnification map.  The source was carried for 250 pixels along part
of path~I, giving 251 brightness values.  Gaussian noise at the level
of 4\% of the current brightness was then added to give the mock
lightcurve.

\subsection{Model-fitting old results}

Note that the mock lightcurve has three nuisance parameters, namely
the brightness normalisation and the start- and endpoints on path~I.
The two latter can also be thought of as the location of the caustic
with respect to the starting point and the speed of the source.

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/cc_forward.eps}
\caption{\label{fig:cc_forward} Mock lightcurve fitted with the
  correct source type and caustic.  The upper panel shows the mock
  data points, and running through them is a seven-parameter best fit:
  these being the crescent parameters $(R_p,R_n,a,b)$ and additionally
  the start and end times and the overall normalisation. The curve has
  $\chi^2=243, P=0.32$ indicating a good fit.  The lower panel shows
  the posterior distributions for $r_{1/2}$ (solid histogram) and
  $R_n/R_p$ (dashed histogram).}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/cc_backward.eps}
\caption{\label{fig:cc_backward} Like Figure~\ref{fig:cc_backward},
  except that the fit uses path~II whereas the mock data still come
  from path~I. The curve has $\chi^2=284, P=0.01$ indicating a
  tolerable fit.}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/gc_forward.eps}
\caption{\label{fig:gc_forward} Like Figure~\ref{fig:cc_backward}, but
  showing the best fit to a Gaussian source.  The histogram shows the
  posterior distribution of $r_{1/2}$ (there is no $R_n$).  The curve
  has $\chi^2=344, P=3\times10^{-6}$ indicating rejection of the
  Gaussian source.}
\end{figure}

@@
